---
title: "Impact of dysfunctional ubiquitination in response to cancer immunotherapy"
subtitle: <center> Somatic mutations analysis - Expression Analysis </center>
author: "Eric Bautista Farrerons (s212514)"
date: '`r paste("First created on May 2024. Updated on ", format(Sys.Date(), "%d %B %Y"))`'
output:
  html_document:
    css: style.css
    code_folding: hide
    fig_caption: yes
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: 4
    number_sections: true
---

# Data pre-processing

This document is to analyse the Expression of the ubiquitin-related genes, immunoproteasome, HLA molecules, etc. and see if there is a pattern that distinguishes responders from non-responders.

```{r, message=FALSE}
### Load packages, functions and paths
source(file = "../01_install.R")
source(file = "../02_functions.R")
source(file = "../03_paths.R")

### Define paths
current_dir <- getwd()
data_path <- file.path(current_dir,
                       "../../data/_raw")
results_pc_path <- file.path(current_dir,
                              "../../results/patient_characteristics/output")
results_fip_path <- file.path(current_dir,
                              "../../results/somatic_mutation_analysis/functional_impact_prediction/output")
results_ubi_db_path <- file.path(current_dir, 
                                 "../../results/somatic_mutation_analysis/ubiquitin_custom_GS/output")
results_ubfora_path <- file.path(current_dir, 
                                 "../../results/somatic_mutation_analysis/ubiquitin_fora/output")
results_protein_seqs_path <- file.path(current_dir, 
                                 "../../results/somatic_mutation_analysis/protein_seqs/output")
results_cellular_loc_path <- file.path(current_dir, 
                                 "../../results/somatic_mutation_analysis/cellular_loc/output")
results_pfam_path <- file.path(current_dir, 
                                 "../../results/somatic_mutation_analysis/pfam_analysis/output")

results_expression_path <- file.path(current_dir, 
                                 "../../results/expression_analysis/output")

# Create the folder if it does not exist
if(!file.exists(results_expression_path)) {
  dir.create(results_expression_path, recursive = TRUE)
}

figures_expression_path <- file.path(current_dir, 
                                 "../../results/expression_analysis/first_analysis/figures")

# Create the folder if it does not exist
if(!file.exists(figures_expression_path)) {
  dir.create(figures_expression_path, recursive = TRUE)
}

### Read data & wrangle
# Clincal metadata
clinical_data_path <- file.path(results_pc_path,
                           "clinical_metadata.csv")
clinical_df <- read.csv(clinical_data_path, header = TRUE)[ , -1]

# Save into RDS file
clinical_metadata_path <- file.path(results_expression_path, 
                      "clinical_metadata.rds")
saveRDS(clinical_df, file = clinical_metadata_path)

# TPM matrix
mm909_path <- file.path(data_path,
                        "MM909_data.RData")
load(mm909_path)
tpm_matrix <- tpm_ensg

# Get hugo_symbols
hugo_symbols_tpm <- rownames(tpm_matrix)

## Filtering out patients with missing data from rna_seq_raw and clinical_df
tpm_matrix_columns <- names(tpm_matrix)

# Subset clinical_df
clinical_df <- clinical_df %>% 
  dplyr::filter(Sample.ID %in% tpm_matrix_columns)

# Rename Sample.ID column
clinical_df <- clinical_df %>% 
  dplyr::rename(SampleID = Sample.ID,
                Response = Patient.Response,
                TMB = mut_load)

# Add hugo_symbols to df
tpm_matrix <- tpm_matrix %>% 
  dplyr::mutate(hgnc_symbol = hugo_symbols_tpm)
```

```{r, warning=FALSE, message=FALSE}
## Convert Hugo symbols to ensembl gene id

#1 Use grch37
ensembl_ids <- SYMBOLtoENSEMBL(hugo_symbols_tpm)

# Filter ensembl gene ids
ensembl_ids_tpm <- ensembl_ids[str_detect(ensembl_ids, "^ENSG")]
remaining_hugo <- ensembl_ids[!str_detect(ensembl_ids, "^ENSG")]

#2 Use grch38 for remaining
ensembl_ids <- SYMBOLtoENSEMBL38(remaining_hugo)

# Filter ensembl gene ids
ensembl_ids_tpm2 <- ensembl_ids[str_detect(ensembl_ids, "^ENSG")]
remaining_hugo <- ensembl_ids[!str_detect(ensembl_ids, "^ENSG")]

# (tried using biomaRt with grch37 but it couldn't map more hugo symbols to ensembl gene ids)

#3 Use biomaRt with grch38
# Connect to the GO BioMart database
ensembl = biomaRt::useMart(biomart="ENSEMBL_MART_ENSEMBL",
                  host="https://www.ensembl.org",
                  path="/biomart/martservice",
                  dataset="hsapiens_gene_ensembl")

# Retrieve hugo & Ensembl gene ids
hugo2ensembl_biomart_res <- biomaRt::getBM(attributes = c("external_gene_name",
                                                        "ensembl_gene_id"),
                   filters = "external_gene_name",
                   values = remaining_hugo,
                   mart = ensembl)

# Make sure 1 to 1 mapping
hugo2ensembl_biomart_df <- hugo2ensembl_biomart_res %>%
  dplyr::rename(hgnc_symbol = external_gene_name) %>% 
  dplyr::arrange(hgnc_symbol, ensembl_gene_id) %>%
  dplyr::group_by(hgnc_symbol) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup()

#4 Merge results
# Convert character vectors to data.frames
df1 <- data.frame(ensembl_gene_id = ensembl_ids_tpm, 
                  hgnc_symbol = names(ensembl_ids_tpm), 
                  stringsAsFactors = FALSE)
df2 <- data.frame(ensembl_gene_id = ensembl_ids_tpm2, 
                  hgnc_symbol = names(ensembl_ids_tpm2), 
                  stringsAsFactors = FALSE)

# Combine all data.frames
combined_df <- rbind(df1, df2, hugo2ensembl_biomart_df)

# Ensure each hgnc_symbol maps to only one ensembl_gene_id
final_ensembl_tpm <- combined_df %>%
  dplyr::group_by(hgnc_symbol) %>%
  dplyr::summarise(ensembl_gene_id = dplyr::first(ensembl_gene_id)) %>%
  dplyr::ungroup()

# Find out missing hugo symbols & convert aliases
remaining_hugo <- setdiff(hugo_symbols_tpm, final_ensembl_tpm$hgnc_symbol)
corrected_remaining_hugo <- alias2SymbolTable(remaining_hugo, species = "Hs")

# Convert character vector to data.frame
df3 <- data.frame(hgnc_symbol = remaining_hugo,
                  corrected_hugo = corrected_remaining_hugo,
                  stringsAsFactors = FALSE)


# Use biomaRt again
hugo2ensembl_biomart_res <- biomaRt::getBM(attributes = c("external_gene_name",
                                                        "ensembl_gene_id"),
                   filters = "external_gene_name",
                   values = df3$corrected_hugo,
                   mart = ensembl)

# Make sure 1 to 1 mapping
hugo2ensembl_biomart_df <- hugo2ensembl_biomart_res %>%
  dplyr::rename(corrected_hugo = external_gene_name) %>% 
  dplyr::arrange(corrected_hugo, ensembl_gene_id) %>%
  dplyr::group_by(corrected_hugo) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup()

# Merge original hugo symbol
df3 <- df3 %>% 
  dplyr::left_join(hugo2ensembl_biomart_df, by="corrected_hugo") %>% 
  dplyr::distinct()

remaining_hugo <- df3 %>% dplyr::filter(is.na(ensembl_gene_id))

# Manually annotate if I can
remaining_hugo <- remaining_hugo %>% 
  dplyr::mutate(
    ensembl_gene_id = case_when(
      hgnc_symbol == "DDIT4-AS1" ~ "ENSG00000269926",
      hgnc_symbol == "LINC01869" ~ "ENSG00000180279",
      hgnc_symbol == "LINC02310" ~ "ENSG00000258537",
      hgnc_symbol == "LINC02618" ~ "ENSG00000293386",
      hgnc_symbol == "LINC02677" ~ "ENSG00000242147"
    )
  )

# Merge
df3 <- df3 %>% 
  dplyr::left_join(remaining_hugo, by=c("hgnc_symbol", "corrected_hugo")) %>% 
  dplyr::mutate(ensembl_gene_id = coalesce(ensembl_gene_id.y, ensembl_gene_id.x)) %>% # Prefer 'remaining_hugo' values
  dplyr::select(-ensembl_gene_id.x, -ensembl_gene_id.y) %>% # Remove the temporary columns
  dplyr::distinct()
remaining_hugo <- df3 %>% dplyr::filter(is.na(ensembl_gene_id))
# 6 hugo symbols remaining that are deprecated

# Remove NAs
final_df3 <- df3 %>% 
  dplyr::filter(!is.na(ensembl_gene_id)) %>% 
  dplyr::select(-corrected_hugo)

# Merge with final df
final_ensembl_tpm <- rbind(final_ensembl_tpm, final_df3)

# Put ensembl gene id to tpm_matrix
final_tpm_matrix <- tpm_matrix %>% 
  dplyr::left_join(final_ensembl_tpm, by="hgnc_symbol") %>% 
  dplyr::filter(!is.na(ensembl_gene_id))

genes <- final_tpm_matrix$ensembl_gene_id
tpm_matrix <- final_tpm_matrix %>% 
  dplyr::select(-hgnc_symbol, -ensembl_gene_id) %>% 
  as.matrix()

rownames(tpm_matrix) <- genes
tpm_matrix_def <- removeDuplicateGenesDF(tpm_matrix)
```

```{r, message=FALSE, warning=FALSE}
# Merge in a list with names = (raw, norm, fpkm, tpm, Clinical)
input_list_expression <- list(
  "raw" = tpm_matrix_def, 
  "norm" = tpm_matrix_def, 
  "fpkm" = tpm_matrix_def, 
  "tpm" = tpm_matrix_def,
  "cpm" = tpm_matrix_def,
  "Clinical" = clinical_df)

# Save into RDS file
input_list_path <- file.path(results_expression_path, 
                      "input_list_expression.rds")
saveRDS(input_list_expression, file = input_list_path)

# Save TPM matrix only
input_tpm_path <- file.path(results_expression_path, 
                      "tpm_matrix.rds")
saveRDS(tpm_matrix_def, file = input_tpm_path)
```

# Merge output of expression analysis with clinical and ubiquitin data (after running Aimilia's pipeline)

```{r, message=FALSE, warning=FALSE}
biomarkers_out <- readRDS(file.path(results_expression_path, "MM909_MEL_TPM_biomarkers.rds"))

# Add clinical metadata & new response categories
biomarkers_clinical <- biomarkers_out %>% 
  dplyr::left_join(clinical_df, by="SampleID") %>% 
  dplyr::mutate(resp_3_cat = case_when(
    RECIST %in% c("CR", "PR") ~ "R", 
    RECIST %in% c("SD") ~ "SD", 
    RECIST %in% c("PD") ~ "NR", 
    TRUE ~ NA_character_
    ))

# Process ubi info and merge
annot_lof_ub_path <- file.path(results_ubfora_path, 
                      "annot_lof_ub.rds")
annot_lof_ub_df <- readRDS(annot_lof_ub_path)

cat_lof_ub_df <- annot_lof_ub_df %>% 
  dplyr::mutate(ub_category = case_when(
    category %in% c("E1_genes", "E2_genes", "E3_genes") ~ "ubiquitination",
    category %in% c("C19_genes", "dub_genes") ~ "deubiquitination",
    TRUE ~ NA_character_
  ))

# Save RDS
cat_lof_ub_path <- file.path(results_expression_path, 
                      "cat_lof_ub.rds")
saveRDS(cat_lof_ub_df, file = cat_lof_ub_path)

reduced_cat_lof_ub_df <- cat_lof_ub_df %>% 
  dplyr::select(sample_id, ub_category) %>% 
  dplyr::group_by(sample_id) %>%
  dplyr::summarise(
    ubi_lof = as.integer(any(ub_category == "ubiquitination")),
    deubi_lof = as.integer(any(ub_category == "deubiquitination"))
  )

# Add ubiquitin lof info
biomarkers_clinical_ubi <- biomarkers_clinical %>% 
  dplyr::left_join(reduced_cat_lof_ub_df, by=c("SampleID" = "sample_id")) %>% 
  tidyr::replace_na(list(ubi_lof = 0, deubi_lof = 0))

# Add ORA results
fora_results_lof_csv_path <- file.path(results_ubfora_path, "fora_results_ubiquitin_lof.csv")
fora_results_ubiquitin_lof_df <- read.csv(fora_results_lof_csv_path, header = TRUE)[ , -1]

fora_ubi_lof_clean_df <- fora_results_ubiquitin_lof_df %>%
  dplyr::rename(count = overlap) %>% 
  dplyr::select(sample_id, pathway, count, size) %>% 
  dplyr::mutate(gene_ratio = count / size) %>% 
  dplyr::filter(!sample_id %in% c("MM909_20", "MM909_24"))

temp_data_df <- fora_ubi_lof_clean_df %>% 
  dplyr::left_join(biomarkers_clinical_ubi_counts %>% dplyr::select(SampleID, TumorPurity), 
                   by=c("sample_id" = "SampleID"))

# Response df
response_df <- biomarkers_clinical_ubi_counts %>% 
  dplyr::rename(sample_id = SampleID) %>% 
  dplyr::select(sample_id, Response) %>% 
  dplyr::distinct()

# Summarize the ubi counts
summarized_data <- temp_data_df %>%
  dplyr::group_by(sample_id) %>%
  dplyr::summarize(
    ubi_counts = sum(count[pathway %in% c("Ubiquitin-protein ligase family", "Ubiquitin-activating enzymes", "Ubiquitin-conjugating enzymes")]),
    deubi_counts = sum(count[pathway %in% c("Deubiquitinases", "Peptidase family C19")])
  )

# Prepare input full df
biomarkers_clinical_ubi_counts <- biomarkers_clinical_ubi %>% 
  dplyr::rename(IMPRES = IMRES,
                IP_Wu = IP_score) %>% 
  dplyr::left_join(summarized_data, by=c("SampleID" = "sample_id"))

# Save into RDS file
full_df_path <- file.path(results_expression_path, 
                      "biomarkers_clinical_ubi_counts.rds")
saveRDS(biomarkers_clinical_ubi_counts, file = full_df_path)
```

## Full output table

```{r}
# Datatable of output
datatable(biomarkers_clinical_ubi_counts, 
          extensions = c('Buttons', 
                         'FixedColumns'), 
          options = list(
            dom = 'Bfrtip',
            buttons = c('copy', 'excel', 'csv'),
            scrollX=TRUE,
            pageLength=8,
            fixedColumns = list(leftColumns = 2), # Freeze the first 2 columns
            columnDefs = list(list(
              targets = "_all",
              render = JS(
                "function(data, type, row, meta) {",
                "  return data === null ? 'NA' : data;",
                "}"
              )
            ))
          ),
          caption = "Expression analysis of Cancer Immunity Cycle biomarkers merged with clinical metadata and LoF ubi mutations (binary and count)."
        )
```

The numeric biomarkers selected to evaluate Cancer Immunity Cycle are the following:

- STEP1: Release_of_cancer_cell_antigens_S1_TIP
- STEP2: Cancer_antigen_presentation_S2_TIP, APM_Senb, APMS_Thomson, IP_Kalaora, IP_Wu, MIAS
- STEP3: Priming_and_activation_S3_TIP
- STEP4: Trafficking_of_immune_cells_into_tumors_overall_S4_TIP
- STEP5: Infiltration_of_immune_cells_into_tumors_S5_TIP, Cytotoxic_cells_cTME, Dendritic_cells_cTME, B_cells_cTME, T_cells_CD8_cTME, T_cells_CD4_cTME, T_regulatory_cells_cTME, Dysfunction, Exclusion
- STEP6: Recognition_of_cancer_cells_by_T_cells_S6_TIP, IMPRES
- STEP7: Killing_of_cancer_cells_S7_TIP, ImmuneCyt_Davoli, Tinflam_Ayers, Ecotype, TME

# Boxplots

Here I compare APM_Senb with APMS_Thomson and IP_Kalaora with IP_Wu to decide only 1 of each pair since they are measuring the same and keeping both would be redundant.

```{r}
# Loop through the biomarkers and generate a boxplot for each
plots <- list()
for (biomarker in c("APM_Senb", "APMS_Thomson", "IP_Kalaora", "IP_Wu")) {
  p <- create_biomarker_boxplot(biomarkers_clinical_ubi_counts, biomarker)
  plots[[biomarker]] <- p
}

# Print the plots
for (biomarker in c("APM_Senb", "APMS_Thomson", "IP_Kalaora", "IP_Wu")) {
  print(plots[[biomarker]])
}
```

Since none is statistically significant, I chose to keep APM_Senb because it is based on HLA genes and IP_Wu because it takes into account the 3 immunoproteasome subunits.

```{r}
# Selection of numeric biomarkers
biomarkers_numeric_df <- biomarkers_clinical_ubi_counts %>% 
  dplyr::select(SampleID, Response, resp_3_cat, Release_of_cancer_cell_antigens_S1_TIP, Cancer_antigen_presentation_S2_TIP, APM_Senb, IP_Wu, MIAS, Priming_and_activation_S3_TIP, Trafficking_of_immune_cells_into_tumors_overall_S4_TIP, Infiltration_of_immune_cells_into_tumors_S5_TIP, Cytotoxic_cells_cTME, Dendritic_cells_cTME, B_cells_cTME, T_cells_CD8_cTME, T_cells_CD4_cTME, T_regulatory_cells_cTME, Dysfunction, Exclusion, Recognition_of_cancer_cells_by_T_cells_S6_TIP, IMPRES, Killing_of_cancer_cells_S7_TIP, ImmuneCyt_Davoli, Tinflam_Ayers, TumorPurity)
biomarkers_numeric_vec <- colnames(biomarkers_numeric_df)[-c(1, 2, 3)]

# Selection of categorical biomarkers
biomarkers_categorical_df <- biomarkers_clinical_ubi_counts %>% 
  dplyr::select(SampleID, Response, resp_3_cat, Ecotype, TME) %>% 
  dplyr::mutate(Response = as.factor(Response),
                resp_3_cat = as.factor(resp_3_cat),
                TME = as.factor(TME),
                Ecotype = as.factor(Ecotype))
biomarkers_categorical_vec <- colnames(biomarkers_categorical_df)[-c(1,2,3)]

# Selection of ubiquitin features
biomarkers_ubiquitin_df <- biomarkers_clinical_ubi_counts %>% 
  dplyr::select(SampleID, Response, resp_3_cat, ubi_lof, deubi_lof, ubi_counts, deubi_counts)
biomarkers_ubiquitin_vec <- colnames(biomarkers_categorical_df)[-c(1,2,3)]

# Individual genes pre-processing
hugo_ensembl_ub <- annot_lof_ub_df %>% 
  dplyr::select(corrected_hugo_symbol, ensembl_id) %>% 
  dplyr::distinct()

ensembl_ub_vec <- hugo_ensembl_ub %>% 
  dplyr::pull(ensembl_id)

## Ubiquitin-related genes with LoF in my data
hugo_ub_vec <- hugo_ensembl_ub %>% 
  dplyr::pull(corrected_hugo_symbol)

## Save into RDS file
ensembl_ub_path <- file.path(results_expression_path, 
                      "ensembl_ub_vector.rds")
saveRDS(ensembl_ub_vec, file = ensembl_ub_path)
hugo_ub_path <- file.path(results_expression_path, 
                      "hugo_ub_vector.rds")
saveRDS(hugo_ub_vec, file = hugo_ub_path)

## Proteasome subunit HGNC symbols
proteas_vec <- c("PSMB5", "PSMB6", "PSMB7", "PSMB8", "PSMB9", "PSMB10")

# Selection of ubiquitin genes
biomarkers_ubi_exp_df <- biomarkers_clinical_ubi_counts %>% 
  dplyr::select(SampleID, Response, resp_3_cat, one_of(hugo_ub_vec))
biomarkers_ubi_exp_vec <- colnames(biomarkers_ubi_exp_df)[-c(1,2,3)]
  
# Selection of proteasome genes
biomarkers_proteas_exp_df <- biomarkers_clinical_ubi_counts %>% 
  dplyr::select(SampleID, Response, resp_3_cat, one_of(proteas_vec))
biomarkers_proteas_exp_vec <- colnames(biomarkers_proteas_exp_df)[-c(1,2,3)]
```

# Compute correlation matrix

```{r, fig.height=10, fig.width=10}
# Fig15

# Filter the data
df2cor <- biomarkers_numeric_df %>% 
  dplyr::select(-SampleID, -Response, -resp_3_cat)

# Compute the correlation matrix
cor_matrix <- cor(df2cor, use = "pairwise.complete.obs", method = "pearson")

# Compute Euclidean distance from the correlation matrix
dist_matrix <- dist(cor_matrix, method = "euclidean")

# Perform hierarchical clustering
hc <- hclust(dist_matrix, method = "complete")

# Order the correlation matrix according to the clustering
ordered_cor_matrix <- cor_matrix[hc$order, hc$order]

# Visualize the correlation matrix with clustering and reduced text size
png(file.path(figures_expression_path, "biomarkers_correlation.png"), width = 1000, height = 800)
corrplot(ordered_cor_matrix, method = "color", type = "upper", order = "original", 
         tl.col = "black", tl.srt = 45, addCoef.col = "black", number.cex = 0.65, tl.cex = 0.8)
dev.off()
```

# Tile plot biomarkers + ubi info

Selected pheatmap_biomarkers_split_response_ubi_binary for thesis, but can uncomment to see other options

```{r, fig.width=8, fig.height=6}
# Fig17

# # pheatmap_biomarkers_both_response_ubi_binary
# p1 <- pheatmap_biomarkers(biomarkers_clinical_ubi_counts, biomarkers_numeric_vec, biomarkers_categorical_vec, response_col = "Response", ubi_counts = FALSE, split_response = FALSE)
# 
# # Save the plot using the png function
# plot_path <- file.path(figures_expression_path, "pheatmap_biomarkers_both_response_ubi_binary.png")
# png(filename = plot_path, width = 8, height = 6, units = "in", res = 300)
# print(p1)
# dev.off()

# pheatmap_biomarkers_split_response_ubi_binary
p2 <- pheatmap_biomarkers(biomarkers_clinical_ubi_counts, biomarkers_numeric_vec, biomarkers_categorical_vec, response_col = "Response", ubi_counts = FALSE, split_response = TRUE)

# Save the ggplot
plot_path <- file.path(figures_expression_path,
                      "pheatmap_biomarkers_split_response_ubi_binary_FAKE.png")
ggsave(plot_path, p2, width = 24, height = 10, dpi = 300)

# # pheatmap_biomarkers_both_response_ubi_counts
# p3 <- pheatmap_biomarkers(biomarkers_clinical_ubi_counts, biomarkers_numeric_vec, biomarkers_categorical_vec, response_col = "Response", ubi_counts = TRUE, split_response = FALSE)
# 
# # Save the plot using the png function
# plot_path <- file.path(figures_expression_path, "pheatmap_biomarkers_both_response_ubi_counts.png")
# png(filename = plot_path, width = 10, height = 6, units = "in", res = 300)
# print(p3)
# dev.off()
# 
# # pheatmap_biomarkers_split_response_ubi_counts
# p4 <- pheatmap_biomarkers(biomarkers_clinical_ubi_counts, biomarkers_numeric_vec, biomarkers_categorical_vec, response_col = "Response", ubi_counts = TRUE, split_response = TRUE)
# 
# # Save the ggplot
# plot_path <- file.path(figures_expression_path,
#                       "pheatmap_biomarkers_split_response_ubi_counts.png")
# ggsave(plot_path, p4, width = 16, height = 10, dpi = 300)
# 
# # pheatmap_biomarkers_both_response_ubi_binary
# p5 <- pheatmap_biomarkers(biomarkers_clinical_ubi_counts, biomarkers_numeric_vec, biomarkers_categorical_vec, response_col = "resp_3_cat", ubi_counts = FALSE, split_response = FALSE)
# 
# # Save the plot using the png function
# plot_path <- file.path(figures_expression_path, "pheatmap_biomarkers_three_response_ubi_binary.png")
# png(filename = plot_path, width = 8, height = 6, units = "in", res = 300)
# print(p5)
# dev.off()
# 
# # pheatmap_biomarkers_split_response_ubi_binary
# p6 <- pheatmap_biomarkers(biomarkers_clinical_ubi_counts, biomarkers_numeric_vec, biomarkers_categorical_vec, response_col = "resp_3_cat", ubi_counts = FALSE, split_response = TRUE)
# 
# # Save the ggplot
# plot_path <- file.path(figures_expression_path,
#                       "pheatmap_biomarkers_three_split_response_ubi_binary.png")
# ggsave(plot_path, p6, width = 24, height = 10, dpi = 300)
# 
# # pheatmap_biomarkers_both_response_ubi_counts
# p7 <- pheatmap_biomarkers(biomarkers_clinical_ubi_counts, biomarkers_numeric_vec, biomarkers_categorical_vec, response_col = "resp_3_cat", ubi_counts = TRUE, split_response = FALSE)
# 
# # Save the plot using the png function
# plot_path <- file.path(figures_expression_path, "pheatmap_biomarkers_three_response_ubi_counts.png")
# png(filename = plot_path, width = 10, height = 6, units = "in", res = 300)
# print(p7)
# dev.off()
# 
# # pheatmap_biomarkers_split_response_ubi_counts
# p8 <- pheatmap_biomarkers(biomarkers_clinical_ubi_counts, biomarkers_numeric_vec, biomarkers_categorical_vec, response_col = "resp_3_cat", ubi_counts = TRUE, split_response = TRUE)
# 
# # Save the ggplot
# plot_path <- file.path(figures_expression_path,
#                       "pheatmap_biomarkers_three_split_response_ubi_counts.png")
# ggsave(plot_path, p8, width = 24, height = 10, dpi = 300)
```

## Ubiquitin LoF and expression plot

```{r, fig.height=14, fig.width=8}
# Fig14

# Normalize biomarkers expression
ub_expression <- biomarkers_ubi_exp_df
ub_expression_vec <- biomarkers_ubi_exp_vec

## Select all columns except the first two for normalization
columns_to_normalize <- colnames(ub_expression)[-(1:3)]

## Perform min-max normalization on the selected columns
ub_expression[columns_to_normalize] <- lapply(ub_expression[columns_to_normalize], function(x) (x - min(x)) / (max(x) - min(x)))

# Pivot long
ub_expression_long <- ub_expression %>% 
  tidyr::pivot_longer(cols = -c(SampleID, Response, resp_3_cat),
                      names_to = "corrected_hugo_symbol",
                      values_to = "expression")

gene_ub_cat <- cat_lof_ub_df %>% 
  dplyr::select(corrected_hugo_symbol, category, ub_category) %>% 
  dplyr::distinct()

correct_hugo <- cat_lof_ub_df %>% 
  dplyr::select(hugo_symbol, corrected_hugo_symbol) %>% 
  dplyr::distinct()

# Identify genes with both "C19_genes" and another category within "deubiquitination"
genes_with_multiple_categories <- gene_ub_cat %>%
  dplyr::filter(ub_category == "deubiquitination") %>%
  group_by(corrected_hugo_symbol) %>%
  dplyr::filter(n_distinct(category) > 1) %>%
  pull(corrected_hugo_symbol) %>%
  unique()

# Filter the data frame to remove "C19_genes" for those genes only
gene_ub_cat <- gene_ub_cat %>%
  dplyr::filter(!(corrected_hugo_symbol %in% genes_with_multiple_categories & ub_category == "deubiquitination" & category == "C19_genes"))

ub_lof_expression <- ub_expression_long %>% 
  dplyr::left_join(gene_ub_cat, by="corrected_hugo_symbol")

lof_ub_binary_path <- file.path(results_ubi_db_path, 
                      "lof_ub_binary.rds")
lof_ub_binary <- readRDS(lof_ub_binary_path)

lof_ub_binary_filtered <- lof_ub_binary %>% 
  dplyr::filter(rowSums(dplyr::select(., -ensembl_id, -hugo_symbol)) != 0) %>% 
  dplyr::left_join(correct_hugo, by="hugo_symbol") %>% 
  dplyr::select(-ensembl_id, -hugo_symbol) %>% 
  dplyr::select(corrected_hugo_symbol, everything())

# Pivot lof_ub_binary_filtered to long format
lof_ub_long <- lof_ub_binary_filtered %>%
  pivot_longer(cols = -corrected_hugo_symbol, 
               names_to = "SampleID", 
               values_to = "ubi_lof")

# Join the data frames
ub_lof_expression_presence <- ub_lof_expression %>%
  dplyr::left_join(lof_ub_long, by = c("SampleID", "corrected_hugo_symbol")) %>% 
  dplyr::mutate(ubi_lof = as.character(ubi_lof)) %>%
  dplyr::mutate(ubi_lof = ifelse(is.na(ubi_lof), "0", ubi_lof))

# Calculate the mean expression for each SampleID
mean_expression <- ub_lof_expression_presence %>%
  group_by(SampleID) %>%
  summarise(mean_expression = mean(expression, na.rm = TRUE)) %>%
  arrange(desc(mean_expression))

# Reorder SampleID based on mean expression
ub_lof_expression_presence$SampleID <- factor(ub_lof_expression_presence$SampleID, levels = mean_expression$SampleID)

# Response plot
response_plot <- ggplot(ub_lof_expression_presence, aes(x = SampleID, y = "1", fill = Response)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_manual(values = c("R" = "#009E73", "NR" = "#D55E00"),
                     labels = c("R" = "Responders", "NR" = "Non-responders")) +
  labs(x = NULL, 
       y = NULL,
       fill = "Patient Response",
       title = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "right",
        legend.spacing.y = unit(0.1, 'cm'),
        legend.background = element_rect(fill = "white", color = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "white", color = NA),
        plot.background = element_rect(fill = "white", color = NA),
        plot.margin = margin(0, 0, 0, 0, "pt"))

# Plot using gradient for expression
plot <- ggplot(ub_lof_expression_presence, aes(x = SampleID, y = corrected_hugo_symbol, fill = expression)) +
  geom_tile(color = "black", linewidth = 0.5, na.rm = FALSE) +
  scale_fill_gradient2(low = "#F0E442", mid = "#009E73", high = "#0072B2", midpoint = 0.5, na.value = "#555555") +
  geom_point(data = ub_lof_expression_presence %>% dplyr::filter(ubi_lof == "1"),
             aes(shape = "Present"), color = "#CC79A7", size = 3, show.legend = TRUE) +
  geom_point(data = ub_lof_expression_presence, aes(size = "NA"), shape = NA, color = "#555555")+
  guides(size=guide_legend("Missing expression", override.aes=list(shape=15, size = 6))) +
  labs(x = "Sample ID", y = "Genes with LoF mutation", fill = "Expression Level", shape = "LoF mutation") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "right",
        legend.spacing.y = unit(0.1, 'cm'),
        legend.background = element_rect(fill = "white", color = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "white", color = NA),
        plot.background = element_rect(fill = "white", colour = NA),
        plot.margin = margin(0, 0, 0, 0, "pt"))

# Extract legends
legend0 <- get_legend(response_plot)
legend1 <- get_legend(plot)

# Combine plots without legends
combined_plot <- response_plot + plot + plot_layout(nrow = 2, ncol = 1, heights = c(0.25, 14)) & theme(legend.position = "none")

# Combine legends
combined_legends <- plot_grid(as_ggplot(legend0) + theme(plot.margin = unit(c(0, 0, 0, 0), "pt")),
                              as_ggplot(legend1) + theme(plot.margin = unit(c(20, 0, 0, 0), "pt")),
                              ncol = 2, align = "v", rel_widths = c(1, 1))

# Combine plots and legends
final_plot <- plot_grid(combined_plot, combined_legends, ncol = 2, rel_widths = c(4, 2), rel_heights = c(20, 1)) +
  theme(plot.background = element_rect(fill = "white", color = NA))

# Print the final plot
print(final_plot)

# Save the ggplot
plot_path <- file.path(figures_expression_path,
                      "ubiquitin_lof_presence_and_expression.png")
ggsave(plot_path, final_plot, width = 10, height = 14, dpi = 300)
```

# Explore individual gene expression profiles

## Ubiquitin gene expression table

```{r}
datatable(biomarkers_ubi_exp_df, 
          extensions = c('Buttons', 
                         'FixedColumns'), 
          options = list(
            dom = 'Bfrtip',
            buttons = c('copy', 'excel', 'csv'),
            scrollX=TRUE,
            pageLength=8,
            fixedColumns = list(leftColumns = 2), # Freeze the first 2 columns
            columnDefs = list(list(
              targets = "_all",
              render = JS(
                "function(data, type, row, meta) {",
                "  return data === null ? 'NA' : data;",
                "}"
              )
            ))
          ),
          caption = "Expression analysis of Ubiquitin-related genes with LoF in samples."
        )
```

## Proteasome subunit expression table

```{r}
datatable(biomarkers_proteas_exp_df, 
          extensions = c('Buttons', 
                         'FixedColumns'), 
          options = list(
            dom = 'Bfrtip',
            buttons = c('copy', 'excel', 'csv'),
            scrollX=TRUE,
            pageLength=8,
            fixedColumns = list(leftColumns = 2), # Freeze the first 2 columns
            columnDefs = list(list(
              targets = "_all",
              render = JS(
                "function(data, type, row, meta) {",
                "  return data === null ? 'NA' : data;",
                "}"
              )
            ))
          ),
          caption = "Expression analysis of Proteasome subunits."
        )
```

## Boxplots of proteasome genes

```{r, fig.width=12, fig.height=8}
#Fig13

proteas_boxplot_path <- file.path(figures_expression_path, "proteasome_boxplot.png")
proteas_boxplot <- create_proteas_boxplot(biomarkers_proteas_exp_df, biomarkers_proteas_exp_vec)

ggsave(proteas_boxplot_path, proteas_boxplot, width = 12, height = 8, dpi = 300)

print(proteas_boxplot)
```

# Combine Tumor Purity with ORA info

```{r}
# Prepare data
fora_results_lof_csv_path <- file.path(results_ubfora_path, "fora_results_ubiquitin_lof.csv")
fora_results_ubiquitin_lof_df <- read.csv(fora_results_lof_csv_path, header = TRUE)[ , -1]

fora_ubi_lof_clean_df <- fora_results_ubiquitin_lof_df %>%
  dplyr::rename(count = overlap) %>% 
  dplyr::select(sample_id, pathway, count, size) %>% 
  dplyr::mutate(gene_ratio = count / size) %>% 
  dplyr::filter(!sample_id %in% c("MM909_20", "MM909_24"))

temp_data_df <- fora_ubi_lof_clean_df %>% 
  dplyr::left_join(biomarkers_clinical_ubi_counts %>% dplyr::select(SampleID, TumorPurity), 
                   by=c("sample_id" = "SampleID"))

response_df <- biomarkers_clinical_ubi_counts %>% 
  dplyr::rename(sample_id = SampleID) %>% 
  dplyr::select(sample_id, Response) %>% 
  dplyr::distinct()

prepared_data_df <- temp_data_df %>% 
  dplyr::left_join(response_df)
```
## Linear regression plot (TumorPurity & Ubi LoF count)

```{r}
# Fig16

prepared_data_df <- temp_data_df %>% 
  dplyr::left_join(response_df)

# Calculate regression statistics
regression_model <- lm(count ~ TumorPurity + Response, data = prepared_data_df)
regression_summary <- summary(regression_model)

# Extract the p-value for Intercept
p_value <- broom::tidy(regression_model) %>%
  dplyr::filter(term == "(Intercept)") %>%
  pull(p.value)

# Create scatter plot with regression line
scatter_plot <- ggplot(prepared_data_df, aes(x = TumorPurity, y = count, color = Response)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("R" = "#009E73", "NR" = "#D55E00"),
                     labels = c("R" = "Responders", "NR" = "Non-responders")) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  labs(title = NULL,
       x = "Tumor Purity",
       y = "nº of LoF mutations",
       color = "Patient Response") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "white", colour = "black"),
        plot.background = element_rect(fill = "white", colour = NA)) +
  annotate("text", x = Inf, y = Inf, label = paste("p-value:", format(p_value, digits = 3)),
           hjust = 1.5, vjust = 25, size = 4, color = "black")

# ggsave
ggsave(file.path(figures_expression_path, "regression_tp_ubi.png"), scatter_plot, width = 12, height = 8, dpi = 300)

# Print the scatter plot
print(scatter_plot)
```

Looks like it is not correlated significatively (p>0.05).

# Session Info

```{r}
sessionInfo()
```
